High-resolution ALMA observations of V4046\,Sgr: a circumbinary disc with a thin ring

### 2021-05-30 Draft_V4046Sgr_mnras.tex
Updates from Overleaf

The multi-frequency data can be interpreted in terms of a physical structure using radiative transfer predictions, for which we used the \textsc{radmc3d} package \citep[version 2.0,][]{Dullemond_2012}. The general framework of the parametric model that we developed is similar to that in \citet{2018MNRAS.477.5104C} for DoAr\,44, and the initial model values were inspired from those in \citet{Rosenfeld_2013}, \citet{Ru_z_Rodr_guez_2019} and \citet{2019ApJ...882..160Q}. Computing a high-resolution radiative transfer model that succeeds in the reproduction of multi-frequency imaging and the SED, is finding a solution to a highly degenerative problem, so obtaining an ideal model with a full parameter exploration requires a level of computation and time that exceeds our capabilities and the scope of this paper. Consequently, our approach was to find through trial and error a set of values for the parameters that closely fit the available data, and then, improve this fit by implementing one dimensional chi squared optimizations for some key parameters (more on this in sec.~\ref{sec:results}). The final structure of the parametric model is summarised in Fig.~\ref{fig:profiles}, where we show on the top panel surface density profiles for gas and the dust grains population, and in the bottom panel the respective aspect ratio profile for each specie.

@@

Although both ALMA and SPHERE--IRDIS images display two-ringed morphologies, we propose a three-ringed structure plus an inner disc to reproduce the observations. This bold decision is due the major fit improvement in the SED and the polarize image (more on this in sec.~\ref{sec:results}). We separate the disc into four individual regions: the inner disc, Ring5, Ring13, and Ring24. The combined gas surface density profile is then given by:
\begin{equation}
  \Sigma_{\mathrm{gas}}(r) = \Sigma_{\mathrm{inner\,disc}} + \Sigma_{\mathrm{Ring5}} + \Sigma_{\mathrm{Ring13}} + \Sigma_{\mathrm{Ring24}}
\end{equation}

The inner disc model follows a power-law function defined by
\begin{equation}
  \Sigma_{\mathrm{inner\,disc}}(r) =\Sigma_\mathrm{c} \left(\frac{r}{R_\mathrm{c}}\right)^{-\gamma}  \, \mathrm{exp}\left[-\left(\frac{r}{R_\mathrm{c}}\right)^{2-\gamma}\right],
\end{equation}
where $R_c$ is a characteristic radius and $\gamma$ is the surface density power-law index. We used $R_c$ = 16\,au, $\Sigma_\mathrm{c} =3.3\times10^{-4}$\,$\mathrm{g\,cm^{-3}}$ and a fixed $\gamma$ = 1 as it is a typical value for discs \citep{Andrews_2009,Andrews_2010}. The gas in our model extends from $R_{sub}$ = 0.3\,au outwards, consistent with the inner edge radius inferred from the SED data (See sec.~\ref{subsec:SED}).

@@

For the reproduction of central emission we introduce a special dust population that is distributed only on a very close Gaussian ring parameterized by $\mu_{\mathrm{central\,blob}}=1.1$\,au, $\sigma_{\mathrm{central\,blob}}=0.4$\,au and $\Sigma_{\circ,\mathrm{central\,blob}}=58.8\,\mathrm{g\,cm^{-3}}$, and following the argument of \citet{Francis_2020}, it goes in past the sublimation radius until it reaches the grid boundary. This dust population has grains with radii ranging from 0.8 to 10\,mm, same upper limit as the large dust but is depleted of small grains. This different size range is a way to avoid creating excess NIR in the SED and to be consistent with previous mass estimates (see Appendix~\ref{sec:Appendix}). As the ring is extremely proximate to the binary system the dust will not have any traces of ice, so we used a composition mix of 70\% silicate and 30\% graphite.

@@

As the radial profiles obtained from the simulated images of the model closely resemble those deduced from the observations (Fig.~\ref{fig:radprofiles}), we can assume that the model provides a possible approximation of the disc structure, including the dimensions of Ring13. The FWHM of the gas in Ring13 in the RT model corresponds to $w_\mathrm{g} = 5.18^{+0.32}_{-0.20}$\,au, but the  large dust grains have to be radially converged to explain the ALMA image. Accordingly, taking the parametric model values for the large dust in Ring13, it has a FWHM of 2.00\,au, a radius of 13.22\,au, and a scale height FWHM of the millimetre-sized dust of 0.25\,au. The total dust mass of Ring13 would be about 0.7\,M$_{\earth}$. The model predictions for the millimetre-sized dust population in Ring13 are close to the measurements, with only a 0.07\,au difference in the centroid location of the ring, and a 0.46\,au difference in the width estimation. Since the scale height FWHM of the large grains in Ring13 is 0.25\,au, and that from the ALMA observation we can resolve a width of 2.46$\pm$0.56\,au (see sec.~\ref{subsec:ALMA}), the ring is 9.8$\pm$1.6 times more extended radially than vertically. Looking at the rest of the disc, in turn, our model reproduces the observations of Ring24 with peak intensity at $\sim$30\,au, and breaks at $\sim$36\,au. It contains a total dust mass of $\sim$45\,M$_{\earth}$.%, that stands under the lower limit of 60\,M$_{\earth}$ inferred by \citet{Ru_z_Rodr_guez_2019} from a previous 860\,$\micron$ continuum ALMA image.

@@

The observed SED is compared with the model in Fig.~\ref{fig:SED}. From the similarity with the data, we propose that there has to be a small-grain population close to the stars down to 0.3\,au ($R_{}$), i.e. an inner disc, as it was crucial to fit the 5--10\,$\micron$ range. On the other hand, the observed central and compact signal in the ALMA observation cannot be explained with the hypothesis of thermal emission from large grains in the inner disc. Following the mass estimates in Section~\ref{sec:Observations}, the millimetre-sized dust needed reaches $\sim$18 times larger surface densities than in the gas, resulting in a large excess at 10$\,\micron$ that does not match the data. Meanwhile, the decision of employing the three-ringed structure (i.e. add Ring5 to the observed structure) relies on the fact that the SED needed a ring at a radius around 4--6\,au to have a proper fit around 10\,$\micron$, otherwise, instead of the distinct bump that is present, there would be a gap.

@@

The parametric model presented here has many degrees of freedom and the RT calculation for each set of parameters is very intensive in computational power and time, so a MCMC optimization or similar methods of parameter exploration are impossible to carry out. As a way to compute a measure of goodness of fit and to try to quantify the model uncertainties, we performed a partial exploration of parameter space. We explore $\chi^2$ space in the vicinity of the model values (obtained by trial and error, see sec.~\ref{sec:model}) for the scale height at $r_\circ$ ($H_\circ$), the width of the gas component of Ring13 $w_{\mathrm{g13}}$, and around the lower limit of the grain sizes in the central blob. The parameter space exploration is shown in Fig.~\ref{fig:chi}. Our model fits the SED and two images, so the total $\chi^2$ value for a given model is composed of the sum of the $\chi^2$ values for each of the three fits.
\begin{equation}
    \chi^2_{\mathrm{total}}=\chi^2_{\mathrm{SED}}+\chi^2_{\mathrm{ALMA}}+\chi^2_{\mathrm{SPHERE}}.
\end{equation}

For $H_\circ$ and $w_{\mathrm{g13}}$, we find that the values of the parameters in our model are almost at a minimum in each one dimensional $\chi^2$ space, so given $\chi^2_\mathrm{m}$ the local minimum value, the uncertainties (up and down) will be those with $\chi^2=\chi^2_\mathrm{m}+1$. As the values are around a local minimum, we can as assume a quadratic equation for the curve fitting.

For the lower grain size limit, we found that although our chosen value is at the minimum, the geometry is more suggestive of a border condition. It is evident that there is a threshold around 500\,$\micron$ where the SED start to deviate far away from the typical value. That point represent where the NIR excess becomes significant.

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{plot_chi_squared_all800.pdf}
        \caption{One dimensional explorations of the chi squared space of our parametric model. Variation of $\chi^2$ in terms of the scale height $H_\circ$ at $r_\circ$=18\,au, the width of the gas component of Ring13 $w_{\mathrm{g13}}$, and around the lower limit of the grain size in the central blob. Each panel displays at the bottom the $\chi^2$ values for the SED and for the two images separately, and on top the total $\chi^2$ value in terms of the parameter value.}
    \label{fig:chi}
\end{figure*}

>>>>

The multi-frequency data can be interpreted in terms of a physical structure using radiative transfer predictions, for which we used the \textsc{radmc3d} package \citep[version 2.0,][]{Dullemond_2012}. The general framework of the parametric model that we developed is similar to that in \citet{2018MNRAS.477.5104C} for DoAr\,44, and the initial model values were inspired from those in \citet{Rosenfeld_2013}, \citet{Ru_z_Rodr_guez_2019} and \citet{2019ApJ...882..160Q}. Computing a high-resolution radiative transfer model that succeeds in the reproduction of multi-frequency imaging and the SED, is finding a solution to a highly degenerative problem, so obtaining an ideal model with a full parameter exploration requires a level of computation and time that exceeds our capabilities and the scope of this paper. Consequently, our approach was to find through trial and error a set of values for the parameters that closely fit the available data, and then, improve this fit by implementing one dimensional chi squared optimizations for some key parameters (more on this in Sec.~\ref{sec:results}). The final structure of the parametric model is summarised in Fig.~\ref{fig:profiles}, where we show on the top panel surface density profiles for gas and the dust grains population, and in the bottom panel the respective aspect ratio profile for each specie.

@@

Although both ALMA and SPHERE--IRDIS images display two-ringed morphologies, we propose a three-ringed structure plus an inner disc to reproduce the observations. This bold decision is due the major fit improvement in the SED and the polarize image (more on this in Sec.~\ref{sec:results}). We separate the disc into four individual regions: the inner disc, Ring5, Ring13, and Ring24. The combined gas surface density profile is then given by:
\begin{equation}
  \Sigma_{\mathrm{gas}}(r) = \Sigma_{\mathrm{inner\,disc}} + \Sigma_{\mathrm{Ring5}} + \Sigma_{\mathrm{Ring13}} + \Sigma_{\mathrm{Ring24}}
\end{equation}

The inner disc model follows a power-law function defined by
\begin{equation}
  \Sigma_{\mathrm{inner\,disc}}(r) =\Sigma_\mathrm{c} \left(\frac{r}{R_\mathrm{c}}\right)^{-\gamma}  \, \mathrm{exp}\left[-\left(\frac{r}{R_\mathrm{c}}\right)^{2-\gamma}\right],
\end{equation}
where $R_c$ is a characteristic radius and $\gamma$ is the surface density power-law index. We used $R_c$ = 16\,au, $\Sigma_\mathrm{c} =3.3\times10^{-4}$\,$\mathrm{g\,cm^{-3}}$ and a fixed $\gamma$ = 1 as it is a typical value for discs \citep{Andrews_2009,Andrews_2010}. The gas in our model extends from $R_{sub}$ = 0.3\,au outwards, consistent with the inner edge radius inferred from the SED data (See Sec.~\ref{subsec:SED}).

@@

For the reproduction of central emission we introduce a special dust population that is distributed only on a very close Gaussian ring parameterized by $\mu_{\mathrm{central\,blob}}=1.1$\,au, $\sigma_{\mathrm{central\,blob}}=0.4$\,au and $\Sigma_{\circ,\mathrm{central\,blob}}=58.8\,\mathrm{g\,cm^{-3}}$, and following the argument of \citet{Francis_2020}, it goes in past the sublimation radius until it reaches the grid boundary. This dust population has grains with radii ranging from 0.8 to 10\,mm, same upper limit as the large dust but is depleted of small grains. This different size range is a way to avoid creating excess NIR in the SED and to be consistent with previous mass estimates (see Appendix~\ref{sec:Appendix} and Sec.~\ref{sec:results}). As the ring is extremely proximate to the binary system the dust will not have any traces of ice, so we used a composition mix of 70\% silicate and 30\% graphite.

@@

As the radial profiles obtained from the simulated images of the model closely resemble those deduced from the observations (Fig.~\ref{fig:radprofiles}), we can assume that the model provides a possible approximation of the disc structure, including the dimensions of Ring13. The FWHM of the gas in Ring13 in the RT model corresponds to $w_\mathrm{g} = 5.18^{+0.32}_{-0.20}$\,au, but the  large dust grains have to be radially converged to explain the ALMA image. Accordingly, taking the parametric model values for the large dust in Ring13, it has a FWHM of 2.00\,au, a radius of 13.22\,au, and a scale height FWHM of the millimetre-sized dust of 0.25\,au. The total dust mass of Ring13 would be about 0.7\,M$_{\earth}$. The model predictions for the millimetre-sized dust population in Ring13 are close to the measurements, with only a 0.07\,au difference in the centroid location of the ring, and a 0.46\,au difference in the width estimation. Since the scale height FWHM of the large grains in Ring13 is 0.25\,au, and that from the ALMA observation we can resolve a width of 2.46$\pm$0.56\,au (see Sec.~\ref{subsec:ALMA}), the ring is 9.8$\pm$1.6 times more extended radially than vertically. Looking at the rest of the disc, in turn, our model reproduces the observations of Ring24 with peak intensity at $\sim$30\,au, and breaks at $\sim$36\,au. It contains a total dust mass of $\sim$45\,M$_{\earth}$.%, that stands under the lower limit of 60\,M$_{\earth}$ inferred by \citet{Ru_z_Rodr_guez_2019} from a previous 860\,$\micron$ continuum ALMA image.

@@

The observed SED is compared with the model in Fig.~\ref{fig:SED}. From the similarity with the data, we propose that there has to be a small-grain population close to the stars down to 0.3\,au ($R_{sub}$), i.e. an inner disc, as it was crucial to fit the 5--10\,$\micron$ range. On the other hand, the observed central and compact signal in the ALMA observation cannot be explained with the hypothesis of thermal emission from large grains in the inner disc. Following the mass estimates in Section~\ref{sec:Observations}, the millimetre-sized dust needed reaches $\sim$18 times larger surface densities than in the gas, resulting in a large excess at 10$\,\micron$ that does not match the data. Meanwhile, the decision of employing the three-ringed structure (i.e. add Ring5 to the observed structure) relies on the fact that the SED needed a ring at a radius around 4--6\,au to have a proper fit around 10\,$\micron$, otherwise, instead of the distinct bump that is present, there would be a gap.

@@

The parametric model presented here has many degrees of freedom and the RT calculation for each set of parameters is very intensive in computational power and time, so a MCMC optimization or similar methods of parameter exploration are impossible to carry out. As a way to compute a measure of goodness of fit and to try to quantify the model uncertainties, we performed a partial exploration of parameter space. We explore $\chi^2$ space in the vicinity of the model values (obtained by trial and error, see Sec.~\ref{sec:model}) for the scale height at $r_\circ$ ($H_\circ$), the width of the gas component of Ring13 $w_{\mathrm{g13}}$, and around the lower limit of the grain sizes in the central blob. The parameter space exploration is shown in Fig.~\ref{fig:chi}. Our model fits the SED and two images, so the total $\chi^2$ value for a given model is composed of the sum of the $\chi^2$ values for each of the three fits.
\begin{equation}
    \chi^2_{\mathrm{total}}=\chi^2_{\mathrm{SED}}+\chi^2_{\mathrm{ALMA}}+\chi^2_{\mathrm{SPHERE}}.
\end{equation}

For $H_\circ$ and $w_{\mathrm{g13}}$, we find that the values of the parameters in our model are at a minimum in each one dimensional $\chi^2$ space, so given $\chi^2_\mathrm{m}$ the local minimum value, the uncertainties (up and down) will be those with $\chi^2=\chi^2_\mathrm{m}+1$. As the values are around a local minimum, we can as assume a quadratic equation for the curve fitting.

For the lower grain size limit, we found that despite our chosen value minimises $\chi^2$, the geometry is more suggestive of a border condition. It is evident that there is a threshold around 500\,$\micron$ where the SED start to deviate far away from the typical value. That point represent where the NIR excess becomes significant.

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{plot_chi_squared_all800.pdf}
        \caption{One dimensional explorations of the chi squared space of our parametric model. We show the variation of $\chi^2$ in terms of the scale height $H_\circ$ at $r_\circ$=18\,au, the width of the gas component of Ring13 $w_{\mathrm{g13}}$, and around the lower limit of the grain size in the close inner ring. Each panel displays at the bottom the $\chi^2$ values for the SED and for the two images separately, and on top the total $\chi^2$ value in terms of the parameter value.}
    \label{fig:chi}
\end{figure*}
